Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-06-13 paper-new/intro.tex
[paper] add outline of remainder of paper to intro

%% \subsubsection{Classical Domain Models}

%%  and gave a denotational
%% model of this calculus using classical domain theory based on
%% $\omega$-CPOs. This semantics has scaled up to an analysis of a
%% dependently typed gradual calculus in \cite{asdf}. This meets our
%% criterion of being a reusable mathematical theory, as general semantic
%% theorems about gradual domains can be developed independent of any
%% particular syntax and then reused in many different denotational
%% models. However, it is widely believed that such classical domain
%% theoretic techniques cannot be extended to model higher-order store, a
%% standard feature of realistic gradually typed languages such as Typed
%% Racket. Thus if we want a reusable mathematical theory of gradual
%% typing that can scale to realistic programming languages, we need to
%% look elsewhere to so-called ``step-indexed'' techniques.

%% \subsubsection{Operational Reduction Proofs}

@@

An alternative approach, which we
investigate in this paper, is provided by \emph{synthetic guarded
domain theory}\cite{birkedal-mogelberg-schwinghammer-stovring2011}.
The techniques of synthetic guarded domain theory allow us to
internalize the step-indexed reasoning normally required in logical
relations proofs of graduality, ultimately allowing us to specify the
logical relation in a manner that looks nearly identical to a typical,
non-step-indexed logical relation. In fact, guarded domain theory goes
further, allowing us to define step-indexed \emph{denotational
semantics} not just step-indexed relations, just as easily as
constructing an ordinary set-theoretic semantics.

In this paper, we develop an adequate denotational semantics that
satisfies graduality and soundness of the equational theory of cast
calculi using the techniques of SGDT.  Our longer-term goal is to
mechanize these proofs in a reusable way in the Guarded Cubical Agda
proof assistant, thereby providing a framework to more easily
and conveniently prove that existing languages satisfy graduality and
have sound equational theories. Moreover, the aim is for designers of
new languages to utilize the framework to facilitate the design of new
provably-correct gradually-typed languages with more complex features.

@@

>>>>

@@

An alternative approach, which we investigate in this paper, is provided by
\emph{synthetic guarded domain
theory}\cite{birkedal-mogelberg-schwinghammer-stovring2011}, abbreviated as
SGDT. The techniques of SGDT allow us to internalize the step-indexed reasoning
normally required in logical relations proofs of graduality, ultimately allowing
us to specify the logical relation in a manner that looks nearly identical to a
typical, non-step-indexed logical relation. In fact, guarded domain theory goes
further, allowing us to define step-indexed \emph{denotational semantics} not
just step-indexed relations, just as easily as constructing an ordinary
set-theoretic semantics.

In this paper, we develop an adequate denotational semantics that satisfies
graduality and soundness of the equational theory of cast calculi using the
techniques of SGDT.

\eric{This is out of date: the mechanization is now an artifact of the current paper.}
Our longer-term goal is to mechanize these proofs in a reusable way in the
Guarded Cubical Agda \cite{veltri-vezzosi2020} proof assistant, thereby
providing a framework to more easily and conveniently prove that existing
languages satisfy graduality and have sound equational theories.

Moreover, the aim is for designers of new languages to utilize the framework to
facilitate the design of new provably-correct gradually-typed languages with
more complex features.

@@

% TODO is it okay to refer to our cast calculus as *the* gradually-typed lambda
% calculus?
\eric{Where should we put the description of the big-step semantics and adequacy
of the semantics? I can put it in the outline of the remainder of the paper, or
I can postpone it until the subsequent section on the syntax of gradual typing.
Adequacy seems to be better introduced there since at that point we have
introduced the notion of term precision.}

The remainder of the paper is organized as follows. In Section
\ref{sec:technical-background}, we provide technical background on synthetic
guarded domain theory and some important details about the specific variant of
SGDT we employ as our ambient language.
%
In Section \ref{sec:GTLC}, we describe the syntactic theory of a standard cast
calculus for gradual typing, which we also refer to as the \emph{gradually-typed
lambda calculus}. We do this to establish a common syntax that will be
interpreted into the models we subsequently construct.
%
In Section \ref{sec:cbpv}, we develop \emph{abstract} categorical
semantics of gradually typed languages with the goal of organizing our
construction of denotational models of gradual typing.
%
In Sections \ref{sec:concrete-term-model} and
\ref{sec:towards-relational-model}, we set out to instantiate the abstract
categorical semantics by constructing a concrete model in a suitable double
category. First, in Section \ref{sec:concrete-term-model}, we give a concrete
model for the \emph{terms} of the gradually-typed lambda calculus. This involves
using the tools of SGDT to solve a domain equation to give the semantics of the
dynamic type. We explain how to \emph{globalize} this model to obtain a model in
the usual (non-guarded) dependent type theory. In the end, we obtain a
\emph{big-step semantics} for closed terms of type $\nat$:
%
\[ \sem{\cdot} \colon (\cdot \vdash M : \nat) \to (\mathbb{N} \to (\sem{nat} + 1)) \] 
%
where a closed term $M$ of type $\nat$ denotes a \emph{function} from
$\mathbb{N}$ to $\sem{nat} + 1$ which given input $i$ returns $\inl\, m$ if $M$
terminates with value $m$ in $i$ or fewer steps, and returns $\inr\, *$ if $M$
fails to terminate in $i$ or fewer steps. This is already a significant result,
as it serves to verify the $\beta$ and $\eta$ laws of the equational theory.
%
Then in Section \ref{sec:towards-relational-model}, we attempt to extend the
methodology of the previous section to give a denotational semantics that models
the relational theory of the gradually-typed lambda calculus in addition to the
terms. Interesting issues arise here arising from the \emph{intensional} nature
of SGDT, that is, the fact that \emph{steps are observable}. We then develop
novel solutions to address these issues. We do not complete the construction of
the concrete relational model in this section. Instead, we choose to reformulate
the definition of abstract categorical model to take into account the solutions
to the issues discussed here.

% Interesting issues arise here, and the novel solutions we employ to
% address these issues inform the definition of the abstract models introduced in
% the subsequent section.

With the lessons from the concrete setting at our disposal, in Section
\ref{sec:abstract-models}, we describe revised categorical models of gradual
typing. We actually introduce two separate notions of model: \emph{intensional},
where the stepping behavior of terms is observable, and \emph{extensional},
where we consider terms that differ only in their stepping behavior to be equal.
The extensional model is what we will ultimately use to interpret the syntax and
relational axioms of gradually-typed languages. The principal benefit of the
notion of intensional model is that it admits more compositional reasoning. It
therefore serves as a convenient intermediate setting that enjoys many of the
benefits of the usual double-categorical models of gradual typing.
% We give the definition of intensional model as a sequence of steps each
% building on the previous. 
We then outline how to build an extensional model in a sequence of steps beginning
with a significantly simpler ``input'' model. The construction proceeds in
steps, first building up an intensional model, and then from there extracting an
extensional model.
%We are able to leverage the compositional reasoning afforded by the intensional
%model to more easily meet the requirements of an extensional model.
This gives us a more tractable way to construct models.
Moreover, the abstract definitions and constructions in the definitions make no
mention of SGDT, so that the places where SGDT techniques are employed are
conveniently limited to the data passed as input to the model construction.
%
%In addition to organizing and facilitating the construction of a model, the abstract constructions are described independently of SGDT
%
%we also isolate the parts of the definition the notion of model without any explicit mention of SGDT.
%
Then in Section \ref{sec:concrete-model} we return to the task of defining a
concrete model. Equipped now with the abstract constructions of the previous
section, we provide only the data needed as an input to the construction. Many
of the relevant definitions have already been introduced in Section
\ref{sec:towards-relational-model} and can be provided to the construction
as-is; the remainder of the necessary data is defined in this section.
%
We end the section by sketching a proof of the \emph{adequacy} of our concrete
model with respect to graduality.
%
Finally, in Section \ref{sec:discussion} we conclude with discussion: comparison
to related work, the mechanization of our concrete model in Guarded Cubical
Agda, and directions for future research.
